---
title: "Final Project"
subtitle: "HPD Crime Data"
author: "Cade Garcia"
institute: "DS303, Chaminade University of Honolulu"
data: "4/25/2025"
format: 
  revealjs:
    theme: "moon"
editor: visual
---

```{r}
#| message: false
#Libraries
library(tidyverse)
library(tidyverse)
library(janitor)
library(lubridate)
library(skimr)
library(ggplot2)
library(plotly)
```

## Intro/Background

::: incremental
-   Data received from the HPD Crime Incidents page on data.honolulu.gov
-   Originally had info meant for record keeping
-   Contains data from Oct 2024 - April 2025 (7 mo)
:::

## Question

::: nonincremental
-   Are some types of crimes more likely to happen at specific times of day?
:::

## Data

```{r}
crime_raw <- read_csv("/Users/Cadeo/Documents/DS303 SP25/DS303_SP25/HPD_Crime_Incidents_20250407.csv") %>%
  clean_names() 

crime <- crime_raw %>%
  mutate(
    datetime = mdy_hms(date), # parse date/time
    hour = hour(datetime), # hour of day (24-hr time)
    wday = wday(datetime, label = TRUE), # get day of week & label it
    is_weekend = wday %in% c("Sat", "Sun") # label weekend vs. weekday (binary)
  ) %>%
  select(incident_num, type, datetime, hour, wday, is_weekend) %>% # narrow down the dataframe to only these columns
  filter(!is.na(hour), !is.na(type)) # drop incomplete rows (those with NAs)

head(crime)
```

6D data with 15,804 rows

## Exploratory Data Analysis 1

::: panel-tabset
### Vis 1

```{r}
crime %>%
  count(type, sort = TRUE) %>%
  ggplot(aes(x = reorder(type, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +  # optional: flips x and y for better readability
  labs(title = "Number of Crimes by Type",
       x = "Crime Type",
       y = "Count") +
  theme_minimal()
```

### Vis 2

```{r}
crime %>%
  count(wday, sort = TRUE) %>%
  mutate(wday = factor(wday, levels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"))) %>%
  ggplot(aes(x = wday, y = n)) +
  geom_col(fill = "steelblue") +
  labs(title = "Number of Crimes by Day",
       x = "Day",
       y = "Count") +
  theme_minimal()

```

### Vis 3

```{r}
crime %>%
  count(is_weekend, sort = TRUE) %>%
  ggplot(aes(x = reorder(is_weekend, n), y = n)) +
  geom_col(fill = "steelblue") +
  labs(title = "Number of crimes week vs. weekend",
       x = "The crime happened on a weekend",
       y = "Count") +
  theme_minimal()
```
:::

## Exploratory Data Analysis 2

```{r}
p <- crime %>% 
  ggplot(aes(x = hour, fill = type)) +
  geom_histogram(binwidth = 1, position = "dodge") +
  labs(title = "Crime Type by Hour of Day", x = "Hour", y = "Count")

ggplotly(p) %>%
    config(modeBarButtonsToRemove = c("zoom2d", "pan2d", "select2d", "lasso2d", "autoScale2d"))
```

## Exploratory Data Analysis 3

```{r}
# looking at whether crime is increasing or decreasing over time
crime_by_day <- crime %>%
  mutate(date = as.Date(datetime)) %>% # convert datetime to date (drops time component)
  count(date) # count number of crimes per day 
```

```{r}
crime_by_day <- crime_by_day[crime_by_day$date != "2025-04-07", ]

ggplot(crime_by_day, aes(date, n)) +
  geom_line() +
  labs(title = "Crime Trend of the Whole Dataset",
       y = "Number of crimes per day",
       x = "Date")
```

## Model Prep

Adds a column that defines violent and non-violent crimes

```{r}
#| echo: true
#| message: true
crime <- crime %>%
  mutate(
    crime_violent = case_when(
      type %in% c("ASSAULT", "ROBBERY", "SEX CRIMES", "WEAPONS") ~ "violent",
      TRUE ~ "non-violent"
    ) %>%
      factor()
  )
```

```{r}
crime$type <- as.factor(crime$type)
crime$wday <- factor(crime$wday, ordered = FALSE)
crime$is_weekend <- as.logical(crime$is_weekend)
crime$is_weekend <- crime$is_weekend == "TRUE"

crime$crime_violent <- as.factor(crime$crime_violent)
```

## Modeling

::: panel-tabset
### Model 1

```{r}
model <- lm(hour ~ type + wday, data = crime)
summary(model)
```

### Model 2

```{r}
model2 <- lm(hour ~ type + wday + is_weekend:type, data = crime)
summary(model2)
```
:::

## Model Comparison

```{r}
anova(model, model2)
```

## Final Thoughts

::: panel-tabset
### Discussion

Focusing on Model 2, The average time for crime is 12.21 (noon). The most significant type of crimes that influenced the model were Fraud, Vandalism, Burglary, Disturbing the peace, and Vehicle theft. The model shows no huge significance or deviation from the average for day of the week. DUIs and Disturbing the Peace crimes did happen more often during the weekends at 6AM and 8AM respectively.

### Limitations

-   Needs more data
-   Yearly data would provide better comparison
:::
